Colonies - Compute Continuums across Platforms

### 2023-04-12 paper_2023/paper.tex
wip

To address these challenges, the paper presents an open-source framework called Colonies\footnote{https://github.com/colonyos/colonies}, which facilitates execution of computational workloads across a diverse range of platforms. Colonies enables development of a \emph{grid-like distributed virtual computer}, called a \emph{Colony}, which can effortlessly integrate with any third-party application (or other workflow management systems).

@@

This paper presents a framework called Colonies, specifically designed to implement a \emph{distributed virtual computer system} that can easily integrate with any third-party system. Colonies is built around a loosely-coupled microservices architecture that separates workflow definitions from implementation and deployment. The primary objective is to establish a platform where monolithic workflows can be decomposed into independent microservices that can easily be integrated with other systems or applications running anywhere on the Internet. The remainder of the paper describes the Colonies framework and how it can be used to create robust and scalable AI systems across platforms.

\section{Related work}
Workflow management has been extensively studied in both academic and industrial settings with numerous approaches \cite{service_wfs, schmitt2022workflow, GarciaRepresa1740746, Ouyang2010, NIKOLOV2021100440, workflow_in_bigdata} proposed to address the challenges in this field. Recently, Apache Airflow \cite{apache_airflow} has become a popular open-source workflow management system for handling data engineering pipelines. Like Colonies, Apache Airflow enables developers to create custom operators and executors that can be integrated with various systems. Additionally, Apache Airflow offers an HTTP API that makes it possible to develop software development kits (SDKs) in various programming languages. However, Apache Airflow does not rely on a queuing system. Instead, it must be integrated with a message broker, such as RabbitMQ \cite{rabbitmq} or Kafka \cite{apache_kafka}, to implement task queues, resulting in a more complex architecture than Colonies. Furthermore, Colonies is based on a distributed microservice architecture that makes it more suitable for DevOps software development. As Colonies is loosely coupled, executors can be implemented and dynamically deployed on the Internet without reconfiguring the workflow engine.

Argo \cite{argowf} is an open source container-native workflow engine for orchestrating parallel jobs on Kubernetes. It can be used for running CI/CD pipelines or compute intensive machine learning or data processing tasks where each job runs as a container. In contrast, Colonies offers a more versatile approach, allowing jobs to be launched within an already started container. As launching new containers on Kubernetes can occasionally be time-consuming, Colonies can deliver higher throughput as the costs of starting new jobs are minimal. This is particularly useful when launching large container images, or workloads (e.g. Julia scripts) that can take a long time to start.

Today, utilization of serverless computing is experiencing a significant growth \cite{cognit}, primarily due to its potential to liberate developers from the burden of managing underlying cloud infrastructures. Attempts are currently being made to implement serverless workflow management systems. For example, OpenWolf \cite{openwolf} is a serverless workflow engine designed to utilize the Function-as-a-Service (FaaS) paradigm for composing complex scientific workflows. It is based on OpenFaaS \cite{openfaas}, which allows functions to run on Kubernetes clusters. 
The serverless workflow project \cite{serverlessworkflows} aims at providing a vendor-neutral workflow DSL. Synapse \cite{synapse} is a workflow management system similar to Argo, but that implements the serverless workflow specification. A similar engine is proposed in \cite{scafe}. 
Colonies can also be used to implement serverless workflow management systems. This will be further discussed in Section \ref{faas}. However, in contrast to previous work, Colonies is designed to be platform independent and does therefore not require Kubernetes. By using a zero-trust security protocol, functions can be securely executed by distributed executors deployed anywhere on the Internet. It is important to point out that Colonies does not provide an infrastructure for function execution. Instead, Colonies primary role is to serve as a platform for coordinating function executions which are carried out by distributed executors.

Currently, microservices are primarily used to implement large-scale web applications or Internet applications requiring high-availability. It has not yet become a prevalent design principle for workload management or implementation of HPC applications. Instead, simple job scripts are commonly used. J. Represa et al. \cite{GarciaRepresa1740746, GarciaRepresa1640771} explore various challenges associated with developing microservice-based workflow management for industrial automation within the context of the Arrowhead project \cite{delsing2017iot}. The authors conclude that microservice-based workflow technologies are viable for industrial applications, particularly due to their inherent flexibility. The primary contribution of this paper is a comprehensive technical description of how to implement a distributed workflow engine based on microservice principles, extending beyond orchestrating microservices to execute automation tasks within one platform. The vision is to create a distributed architecture where microservices can reside anywhere on the Internet and still function as a cohesive unit to execute cross-platform workloads.

@@

\section{The Colonies framework}
\label{sec:headings}
A core concept of the Colonies framework is the notion of \emph{processes}. A process contains meta-information about computations executed by remote computer programs, referred to as \emph{distributed executors}. Specifically, a process contains a definition of a function to be executed, as well as contextual information such as execution status, including the result of the computation. One can think of a process as a digital twin of a real computing process. It is essential to note that a process does not necessarily have to be an operating system process, but can represent any type of computation, e.g. a remote procedure call executed by any kind of software.

@@

\begin{itemize}
\item \textbf{Single responsibility:} Each executor is only accountable for executing specific functions. This makes the system easier to understand, develop, test and maintain.  
\item \textbf{Loosely coupled:} Executors are designed with minimal dependencies on other executors, enabling various software development teams to work independently. For instance, a data engineering team may handle the implementation of pre-processing functions, while a data science team oversees machine learning functions, and another team manages visualization or customer integration etc.
\item \textbf{Scalability:} Executors can be deployed independently, which enables horizontal scaling. This allows for better resource utilization, parallelism, and improved performance. 
\item \textbf{Resilience:} The failure of a single executor does not compromise the entire application or workflow. Executors' isolation from one another contributes to a more resilient and fault-tolerant system. If an executor crashes during execution, the process is automatically reassigned to another executor.
\item \textbf{DevOps and Continuous Integration:} Executors' inherent fault-tolerant design permits changes to individual executors without impacting the entire system. This makes it possible to seamlessly update the system. 
\item \textbf{Technology agnostic:} Executors can be implemented in any programming language, facilitating seamless integration with other applications and systems.
\item \textbf{Decentralized governance:} As the Colonies framework is technology-agnostic, different software development teams can make independent technology and design decisions when developing executors, promoting greater flexibility and adaptability. For example, some executors may be implemented in Rust, while others may use Python to leverage state-of-the-art machine learning frameworks.
\end{itemize}

Although microservices have attractive properties and simplify both development and deployment of executors, they also introduce additional complexity that must be managed by the Colonies servers. To implement a workflow management framework supporting distributed microservices, the following challenges must be addressed:

\begin{itemize}
\item \textbf{Process management:} Colonies must be able to distribute processes among available executors based on their capabilities and current workload. This involves assigning processes to the most appropriate executor and then monitoring process progress.

@@

The Colonies framework uses a combination of technologies to address the aforementioned challenges. A fundamental design principle is statelessness, which makes Colonies simpler, more reliable and scalable, and easier to implement. To ensure reliability and data consistency, a distributed consensus algorithm is needed to handle high-availability and server crashes. The remainder of this section describes how the Colonies framework is implemented. The next discusses the role of queues to support batch jobs and realtime processing, but also how to implement a loosely coupled system.

@@

\subsection{The role of queues as seperation of concerns}
Separation of concerns (SoC) is a design principle to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, HTTP protocols alone are insufficient for handling dynamic environments where components frequently fail or the underlying infrastructure is constantly changing. To address such environments, an alternative mechanism is necessary.

Queues enable software services to communicate indirectly by acting as a buffer between them. Queues make it possible to decouple each executor and make them operate independently, e.g. an executor can be updated or replaced without affecting other executors. Queues also enable asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantly, queues enable load balancing by distributing tasks among multiple executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in different ways, and while message brokers are a common solution, Colonies adopt an alternative strategy and leverage a standard database and query it for tasks to assign to different executors. One key advantage of this approach is that it enables fine-grained process assignments, making it possible to assign specific processes to particular executors. For instance, an executor of the browser type can be limited to only executing processes in web applications. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queues, or provide the ability to pull specific messages out of the queue. Generally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, a database can function as a queue and a query can match any columns thus making it possible to assign specific executors to specific processes.

\subsection{Process tables}
Colonies enables executors to interact with each other by submitting function specifications to the Colonies servers. Once submitted, other executors can connect to the servers to receive process execution assignments. A new process entry is added to the process table database when a function specification is submitted to the Colonies server.

@@

When an executor connects to the Colonies server, the server hangs the incoming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. Note that the Colonies server does not connect to the executors. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, behind firewalls, commercial telco networks, or even in web browser-based applications.

@@

Figure \ref{fig:process_assignment} depicts an executor submitting a function specification that is later assigned to another executor. When registering, executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server then makes sure that the conditions (i.e requirements) part of the function specification matches the capability of an executor.

@@

Table \ref{proctable} shows an example of a process table. The process table also contains a reference to a function specification depicted in Table \ref{functable}. To assign a process to an executor, the Colonies server searches in the process table to find a process matching a waiting executor. By using the \emph{priority time} column to sort the processes, the process table can function as a queue. This can be done in SQL by utilizing the \emph{order by} clause to sort processes according to their submission time. To make it possible to handle priorities, the submission time is adjusted so that higher priority processes are processed before lower priority processes. When a process is submitted, a \emph{priority time} value is calculated and stored in the process table. Equation \ref{eq:pt} shows how the priority time is calculated for a nanosecond timestamp.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.6]{raft.png}
	\caption{High-availability deployment of Colonies.}
	\label{fig:ha_deployment}
\end{figure}

\subsection{A stateless failsafe mechanism}
\label{stateless_error_management}
The primary objective of all Colonies API requests is to alter some state stored in the database or retrieve information from the database. The Colonies framework is designed to be stateless, meaning that the Colonies server does not keep any information in memory between requests. Each request is handled independently without relying on information from previous requests.

@@

\subsubsection{Data consistency and distributed consensus}
Synchronization is essential to prevent data inconsistency and race conditions when accessing shared resources concurrently with multiple threads. However, synchronization can also slow down execution as only one thread can access critical sections at a time. By carefully designing multithreaded applications and employing the right synchronization techniques, it is possible to minimize the performance impact while still ensuring data consistency and correctness.

@@

By using the Raft protocol, Colonies can redirect incoming \emph{assign} requests to the leading Colonies server, thereby ensuring that only one Colonies server replica handles such requests. A new leader is automatically elected in the event that a Colonies server replica fails. Figure \ref{fig:ha_deployment} shows an overview of a high-availability Colonies deployment. In addition, the Raft protocol makes it possible to seamlessly update Colonies software by making it possible to upgrade each Colonies server replica individually. Consequently, making Colonies well-suited for Kubernetes deployments.

@@

When a process is closed, the Colonies server checks its child processes to see if all their parent processes have also completed. If all parents of a child process have finished, the server updates the process table in the database by setting the flag \emph{wait for parents} to false, which allows child processes to be assigned to executors. This operation is also done statelessly when processing the \emph{close} request.

Figure \ref{fig:workflowexec} portrays an execution timeline for a workflow. Upon submission, process \(P_{1}\) is assigned to Executor \(E_{1}\). After being closed, processes \(P_{2}\) and \(P_{3}\) are simultaneously assigned to \(E_{1}\) and \(E_{2}\), allowing for concurrent execution. Lastly, process \(P_{4}\) is assigned to \(E_{2}\).

To be able to generate a DAG, an additional database table is required. Table \ref{deptable} shows an example of a dependency table storing relationships between processes. The table is updated when a workflow is submitted to a Colonies server. As mentioned before, when an executor is assigned a process it obtains exclusive rights to that process within the process table. This exclusivity also means that it is possible for an exector to dynamically add new children processes to the workflow DAG on the fly. As a result, it becomes possible to implement MapReduce \cite{mapreduce} data processing patterns similar to Hadoop \cite{hadoop}.

@@

\subsubsection{Cron}
The Colonies framework is based on a stateless architecture where each request is treated as an independent transaction and is processed without any information from prior requests. Since there is no session data stored in memory, Colonies can easily switch between servers without disruptions or data losses. In a stateful architecture on the other hand, the server must maintain a record of connected clients' state and session data in memory, which can result in complications if a server fails or crashes.

Cron is a time-based job scheduler designed to trigger workflows at predefined intervals. It can for example be used to automate tasks such as routinely fetching data from external systems. To make Colonies robust and easy to implement it is essential that Cron workflows also are managed stateless. This can be achieved by introducing an additional database table to store information about Cron workflow, and let the elected Colonies server leader assume the responsibility of managing Cron workflows through a two-step process.

Firstly, the leading Colonies server calculates a future deadline, indicating when a specific Cron workflow is scheduled to be executed. The leader server then conducts periodic scans of the Cron table. When the current time surpasses the calculated deadline for a particular workflow, a new workflow is automatically submitted. Note that this protocol is completely stateless as no session information is stored in memory between cron scans. In the event that the leader Colonies server experiences a crash, a new leader is simply selected and assumes the responsibility of managing Cron workflows according to the two-step protocol.

@@

Generators facilities integration with third-party systems as it allows for stateless implementation where interactions with Colonies can be done in a fire-and-forget style. Colonies guarantees that all pack requests submitted to a generator are processed correctly, even if a Colonies server is restarted or fails. Without generators, third-party integration software would have to use an alternative method such as a message broker or a database to enqueue input data and periodically empty the queues to generate workflows themselves. This adds complexity to the integration process and may require additional infrastructure components. Generators are a powerful feature of the Colonies framework that can help simplify integration and implement robust data processing pipelines.

Generators can be implemented using a similar method to how Cron workflows are implemented. The elected leader Colonies server takes on the responsibility for scanning a special generator table to decide if a workflow should be triggered. This approach ensures that there is a single point of control for generating workflows and prevents generation of duplicate workflows. If the leader server crashes, a new leader Colonies server is automatically elected, which takes over the responsibility for triggering generators. This ensures that the generation of workflows is not impacted by server failures, and the system can continue to operate smoothly even in case of failures. It can be worth pointing out that since the pack request only adds information to the generator table and does not manipulate any states, it can be processed by any Colonies server without synchronization precautious. This means that any Colonies server in the cluster can handle incoming pack requests, which increases the scalability and fault-tolerance of the system as a whole.

@@

As previously stated, a colony is a collection of distributed executors. Executors part of the same colony trust each other and can thus submit function specifications and get process execution assignments. To implement such a scheme, the Colonies server must be able to verify the identity of the executors and then check their colony membership. This can be accomplished in various ways. One solution is to use public key encryption and assign each executor a pair of keys - a public key and a private key. The public key is openly available and uploaded to the Colonies server, whereas the private key is kept secret and is only used by the executor to sign API request messages. The Colony server can then verify the signature of incoming request messages and use the public keys to determine whether an executor is a member of the colony it is attempting to interact with.

@@

An advantage of using a stateless architecture for workflow management as proposed in the paper is the ability to monitor executor performance to detect any unusual or suspicious behaviors. Because all transactions are recorded in a database, it is possible to track the activity of individual executors and identify any unusual patterns or behaviors. For example, if an executor suddenly starts to consume a lot of resources, it could indicate a performance issue, or it could be a sign of malicious activity, such as a denial-of-service attack. Automatic monitoring can for example be implemented using anomaly detection algorithms and machine learning to grade the performance of each executor. Administrators can then be automatically notified if the grade of an executor exceeds a certain threshold, enabling them to take appropriate actions.

\section{Implementation}
Colonies is implemented in Golang and is publicly available on GitHub\footnote{https://github.com/colonyos/colonies} under the MIT license. It consists of a statically compiled binary that offers a CLI tool for deploying Colonies servers or managing the system. Furthermore, there are several software development kits (SDKs) available for various programming languages, including Golang\footnote{https://github.com/colonyos/colonies/tree/main/pkg/client}, Rust\footnote{https://github.com/colonyos/rust}, Julia\footnote{https://github.com/colonyos/Colonies.jl}, JavaScript\footnote{https://github.com/colonyos/colonies.js}, Python\footnote{https://github.com/colonyos/pycolonies}, and Haskell\footnote{https://github.com/colonyos/haskell}. As an introduction to Colonies, the following section gives a brief overview of the Colonies tools and how to use the Python SDK.

@@

The first step in implementing an executor in Python is to create a private key and identity, followed by registering the executor with a colony. Registering an executor can only be carried out by the colony owner with access to the colony private key. Code in Listening \ref{code:regexecutor} illustrates how to register an executor and which functions a specific executor is capable of running. In the example, an executor of type \emph{helloworld\_executor} is registered to run a function named \emph{helloworld}.

@@

After registration, the executor must connect to the Colonies server and request process assignments of the type \emph{helloworld\_executor}. Note that executor properties, such as the executor type, are not explicitly specified, but rather derived implicitly by the Colonies server from the executor identity, which is derived from the signature generated by the executor's private key. This approach ensures that the executor can only be assigned processes that match its registered capabilities.

@@

This \emph{assign} request is always initiated by the executor using an HTTP-based protocol, allowing executors to be deployed behind firewalls. When receiving the request, the Colonies server hangs the request until a matching process is found or a timer expires. In the example in Listing \ref{code:assign}, the timer is set for 10 seconds, after which an exception is raised, and the executor must reconnect to the server.

When assigned to a process, the executor interprets the metadata stored in the process data structure, including the function name and arguments, and performs some kind of computation such as preprocessing some data or training a neural network. Upon completion, the executor closes the process and sets the output. In the example, the output is set to the string \emph{helloworld}.

@@

\section{Appendix - Colonies operations}

>>>>

To address these challenges, the paper presents an open-source framework called Colonies\footnote{https://github.com/colonyos/colonies}, which facilitates execution of computational workloads across a diverse range of platforms. Colonies enables development of a \emph{decentralized virtual computer system}, called a \emph{Colony}, which can effortlessly integrate with any third-party application or other workflow systems.

@@

This paper presents a framework called Colonies, specifically designed to implement a \emph{distributed virtual computer system} that can easily integrate with any third-party system. Colonies is built around a loosely-coupled microservices architecture that separates workflow definitions from implementation and deployment. The primary objective is to establish a platform where workflows can be decomposed into independent microservices that can easily be integrated with other systems or applications running anywhere on the Internet. The remainder of the paper describes the Colonies framework and how it can be used to create robust and scalable AI systems across platforms.

\section{Related work}
Workflow management has been extensively studied in both academic and industrial settings with numerous approaches \cite{scafe, synapse, service_wfs, schmitt2022workflow, GarciaRepresa1740746, Ouyang2010, NIKOLOV2021100440, workflow_in_bigdata} proposed to address the challenges in this field. Recently, Apache Airflow \cite{apache_airflow} has become a popular open-source workflow management system for handling data engineering pipelines. Like Colonies, Apache Airflow enables developers to create custom operators and executors that can be integrated with various systems. Additionally, Apache Airflow offers an HTTP API that makes it possible to develop Software Development Kits (SDKs) in various programming languages. However, Apache Airflow must be integrated with a message broker, such as RabbitMQ \cite{rabbitmq} or Kafka \cite{apache_kafka} to implement task queues, resulting in a more complex architecture than Colonies. Furthermore, Colonies is based on a distributed microservice architecture that makes it more suitable for DevOps software development. As Colonies is loosely coupled, executors can be implemented and dynamically deployed on the Internet without reconfiguring the workflow engine.

Argo \cite{argowf} is an open source container-native workflow engine for orchestrating parallel jobs on Kubernetes. It can be used for running CI/CD pipelines or compute intensive machine learning or data processing jobs where each job runs as a container. In contrast, Colonies offers a more versatile approach, allowing jobs to be launched within an already started container. As launching new containers on Kubernetes can occasionally be time-consuming, Colonies can deliver higher throughput as the costs of starting new jobs are minimal. This is particularly useful when launching large container images, or workloads (e.g. Julia scripts) that can take a long time to start.

Today, utilization of serverless computing is experiencing a significant growth \cite{cognit}, primarily due to its potential to liberate developers from the burden of managing underlying cloud infrastructures. Attempts are currently being made to implement serverless workflow management systems. For example, OpenWolf \cite{openwolf} is a serverless workflow engine designed to utilize the Function-as-a-Service (FaaS) paradigm for composing complex scientific workflows. It is based on OpenFaaS \cite{openfaas}, which allows functions to run on Kubernetes clusters.

Colonies can also be used to implement serverless workflow management systems. This will be further discussed in Section \ref{faas}. However, in contrast to previous work, Colonies is designed to be platform independent and does therefore not require Kubernetes. By using a zero-trust security protocol, functions can be securely executed by distributed executors deployed anywhere on the Internet. It is important to point out that Colonies does not provide an infrastructure for function execution. Instead, Colonies primary role is to serve as a platform for coordinating function executions which are carried out by distributed executors.

Currently, microservices are primarily used to implement large-scale web applications or Internet applications requiring high-availability. It has not yet become a prevalent design principle for workload management or implementation of HPC applications. Instead, simple job scripts are commonly used. J. Represa et al. \cite{GarciaRepresa1740746, GarciaRepresa1640771} explore various challenges associated with developing microservice-based workflow management for industrial automation within the context of the Arrowhead project \cite{delsing2017iot}. The authors conclude that microservice-based workflow technologies are viable for industrial applications, particularly due to their inherent flexibility. The primary contribution of this paper is a comprehensive technical description of how to implement a distributed workflow engine based on microservice principles, extending beyond orchestrating microservices to execute automation tasks within one platform. The vision is to create a distributed architecture where microservices can reside anywhere on the Internet and still function as a cohesive unit to execute cross-platform workloads.  
o

@@

\section{The Colonies framework}
\label{sec:headings}
A core concept of the Colonies framework is the notion of \emph{processes}. A process contains meta-information about computations executed by remote computer programs, referred to as \emph{distributed executors}. Specifically, a process contains a definition of a function to be executed as well as contextual information such as execution status, including the result of the computation. One can think of a process as a digital twin of a real computing process. It is essential to note that a process does not necessarily have to be an operating system process, but can represent any type of computation, e.g. a remote procedure call executed by any kind of software.

@@

\begin{itemize}
\item \textbf{Single responsibility:} Each executor is only accountable for executing specific functions. This makes the system easier to understand, develop, test and maintain.  
\item \textbf{Loosely coupled:} Executors are designed with minimal dependencies on other executors, enabling various software development teams to work independently. For instance, a data engineering team may handle the implementation of pre-processing functions, while a data science team oversees machine learning functions, and another team manages visualization or customer integration etc.

\item \textbf{Scalability:} Executors can be deployed independently, which enables horizontal scaling. This allows for better resource utilization, parallelism, and improved performance.

\item \textbf{Resilience:} The failure of a single executor does not compromise the entire application or workflow. Executors' isolation from one another contributes to a more resilient and fault-tolerant system. If an executor crashes during execution, the process is automatically reassigned to another executor.

\item \textbf{DevOps and Continuous Integration:} Executors' inherent fault-tolerant design permits changes to individual executors without impacting the entire system. This makes it possible to seamlessly update the system.

\item \textbf{Technology agnostic:} Executors can be implemented in any programming language, facilitating seamless integration with other applications and systems.

\item \textbf{Decentralized governance:} As the Colonies framework is technology-agnostic, different software development teams can make independent technology and design decisions when developing executors, promoting greater flexibility and adaptability. For example, some executors may be implemented in Rust, while others may use Python to leverage state-of-the-art machine learning frameworks.
\end{itemize}

Although microservices have attractive properties and can simplify both development and deployment of executors, they also introduce additional complexity that must be managed by the Colonies framework. To implement a workflow management framework supporting distributed microservices, the following challenges must be addressed:

\begin{itemize}
\item \textbf{Process management:} Colonies must be able to distribute processes among available executors based on their capabilities and current workload. This involves assigning processes to the most appropriate executor and then monitoring execution progress.

@@

The Colonies framework uses a combination of technologies to address the aforementioned challenges. A fundamental design principle is statelessness, which makes Colonies simpler, more reliable and scalable, and easier to implement. To ensure reliability and data consistency, a distributed consensus algorithm is needed to handle high-availability and server crashes. The remainder of this section describes how the Colonies framework is implemented. The next discusses the role of queues to support batch jobs and realtime processing, but also queues roles in implementing loosely coupled systems.

@@

\subsection{The role of queues as seperation of concerns}
Separation of concerns (SoC) is a design principle to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, HTTP APIs alone are insufficient for handling dynamic environments where components frequently fail or the underlying infrastructure is constantly changing. To address such environments, an alternative mechanism is needed.

Queues enable software services to communicate indirectly by acting as a buffer between them. Queues make it possible to decouple executor and make them operate independently, for example an executor can be updated or replaced without affecting other executors. Queues also enable asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantly, queues enable load balancing by distributing tasks among executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in different ways. While message brokers are a common solution, Colonies adopt an alternative strategy and leverage a standard database and query it for processes to assign to different executors. One key advantage of this approach is that it enables fine-grained process assignments, making it possible to assign specific processes to particular executors. For instance, an executor of the browser type can be limited to only executing processes capable of running in web browsers. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queues, or provide the ability to pull specific messages out of the queue. Generally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, a database can function as a queue and a query can match any columns making it possible to assign specific executors to specific processes.

\subsection{Process tables}
Colonies enables executors to interact with each other by submitting function specifications to the Colonies servers. Once submitted, other executors can connect to the servers to receive process execution assignments. A new process entry is automatically added to the process table database when a function specification is submitted to the Colonies server.

@@

When an executor connects to the Colonies server, the server hangs the incoming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. The Colonies server never connects to the executors. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, behind firewalls, commercial telco networks, or even in web browser-based applications.

@@

Figure \ref{fig:process_assignment} depicts an executor submitting a function specification that is later assigned to another executor. When registering, executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server then makes sure that the conditions part of the function specification matches the capability of an executor.

@@

Table \ref{proctable} shows an example of a process table. The process table also contains a reference to a function specification depicted in Table \ref{functable}. To assign a process to an executor, the Colonies server searches in the process table to find a process matching a waiting executor. By using the \emph{priority time} column to sort the processes, the process table can function as a queue. This can be done in SQL by utilizing the \emph{order by} clause to sort processes according to their submission time.

To make it possible to handle priorities, the submission time is adjusted so that higher priority processes are processed before lower priority processes. When a process is submitted, a \emph{priority time} value is calculated and stored in the process table. Equation \ref{eq:pt} shows how the priority time is calculated for a nanosecond timestamp.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.5]{raft.png}
	\caption{High-availability deployment of Colonies.}
	\label{fig:ha_deployment}
\end{figure}

\subsection{A stateless failsafe mechanism}
\label{stateless_error_management}
The primary objective of all Colonies API requests is to alter some state stored in the database or retrieve information from the database. The Colonies framework is designed to be stateless, meaning that Colonies servers do not keep any information in memory between requests. Each request is handled independently without relying on information from previous requests.

@@

\subsubsection{Data consistency and distributed consensus}
Synchronization is essential to prevent data inconsistency and race conditions when accessing shared resources concurrently with multiple threads. However, synchronization can also slow down execution as only one thread can access critical sections at a time. By carefully designing multithreaded applications and using right synchronization techniques, it is possible to minimize the performance impact while still ensuring data consistency and correctness.

@@

By using the Raft protocol, Colonies can redirect incoming \emph{assign} requests to the leading Colonies server, thereby ensuring that only one Colonies server replica handles assign requests. A new leader is automatically elected in the event that a Colonies server replica fails. Figure \ref{fig:ha_deployment} shows an overview of a high-availability Colonies deployment. In addition, the Raft protocol makes it possible to seamlessly update Colonies software by making it possible to upgrade each Colonies server replica individually. Consequently, making Colonies servers well-suited for Kubernetes deployments.

@@

When a process is closed, the Colonies server checks its child processes to see if all their parent processes have also completed. If all parents of a child process have finished, the server updates the process table in the database by setting the flag \emph{wait for parents} to \emph{false}, which allows child processes to be assigned to executors. This operation is also done statelessly when processing the \emph{close} request.

Figure \ref{fig:workflowexec} shows an execution timeline for a workflow. Upon submission, process \(P_{1}\) is assigned to Executor \(E_{1}\). After being closed, processes \(P_{2}\) and \(P_{3}\) are simultaneously assigned to \(E_{1}\) and \(E_{2}\), allowing for concurrent execution. Lastly, process \(P_{4}\) is assigned to \(E_{2}\).

To be able to generate a DAG, an additional database table is required. Table \ref{deptable} shows an example of a dependency table storing relationships between processes. The table is updated when a workflow is submitted to a Colonies server. As mentioned before, when an executor is assigned a process, it obtains exclusive rights to that process within the process table. This exclusivity also means that it is possible for an exector to dynamically add new children processes to the workflow DAG on the fly. As a result, it becomes possible to implement MapReduce \cite{mapreduce} data processing patterns similar to Hadoop \cite{hadoop}.

@@

\subsubsection{Cron}
The Colonies framework is based on a stateless architecture where each request is treated as an independent transaction which is processed without any information from prior requests. Since there is no session data stored in memory, Colonies can easily switch between servers without disruptions or data losses. In a stateful architecture on the other hand, the server must maintain a record of connected clients' state and session data in memory, which can result in complications if a server fails or crashes.

Cron is a time-based job scheduler designed to trigger workflows at predefined intervals. It can for example be used to automate tasks such as routinely fetching data from external systems. To make Colonies robust and easy to implement, it is essential that Cron workflows also are managed stateless. This can be achieved by introducing an additional database table to store information about Cron workflow, and let the elected Colonies server leader assume the responsibility of managing Cron workflows through a two-step process.

Firstly, the leading Colonies server calculates a future deadline, indicating when a specific Cron workflow is scheduled to be executed. The leader server then conducts periodic scans of the Cron table. When the current time surpasses the calculated deadline for a particular workflow, a new workflow is automatically submitted. Note that this protocol is completely stateless as no session information is stored in memory between cron scans. In the event that the leader Colonies server experiences a crash, a new leader is selected and assumes the responsibility of managing Cron workflows according to the two-step protocol.

@@

Generators facilities integration with third-party systems as it allows for stateless implementation where interactions with Colonies can be done in a fire-and-forget style. Colonies guarantees that all pack requests submitted to a generator are processed correctly, even if a Colonies server is restarted or fails. Without this robustness guarantee, third-party integration software would have to use an alternative method such as a message broker or a database to enqueue input data and periodically empty the queues to generate workflows themselves. This adds complexity to the integration process and may require additional infrastructure components. Generators are a powerful feature of the Colonies framework that can help simplify integration and implement robust data processing pipelines.

Generators can be implemented using a similar method to how Cron workflows are implemented. The elected leader Colonies server takes on the responsibility for scanning a special generator table to decide if a workflow should be triggered. This approach ensures that there is a single point of control for generating workflows and prevents generation of duplicate workflows. If the leader server crashes, a new leader Colonies server is automatically elected, which takes over the responsibility for triggering generators. This ensures that the generation of workflows is not impacted by server failures, and the system can continue to operate smoothly even in case of failures. It can be worth pointing out that since the \emph{pack} request only adds information to the generator table and does not manipulate any states, it can be processed by any Colonies server without synchronization precautious. This means that any Colonies server in the cluster can handle incoming \emph{pack} requests, which increases the scalability and fault-tolerance of the system as a whole.

@@

As previously stated, a colony is a collection of distributed executors. Executors part of the same colony trust each other and can thus submit function specifications and get process execution assignments. To implement such a scheme, the Colonies server must be able to verify the identity of the executors and then check their colony membership. This can be accomplished in various ways. One solution is to use public key encryption and assign each executor a pair of keys - a public key and a private key. The public key is openly available and uploaded to the Colonies server, whereas the private key is kept secret and is only used by the executor to sign API request messages. The Colony server can then verify the signature of incoming request messages and use the public keys to determine whether an executor is a member of a certain colony.

@@

An advantage of using a stateless architecture for workflow management as proposed in the paper is the ability to monitor executor performance to detect any unusual or suspicious behaviors. Because the execution history is recorded in a database, it is possible to track the activity of individual executors and identify any unusual patterns. For example, if an executor suddenly starts to consume a lot of resources, it could indicate a performance issue or it could be a sign of malicious activity, such as a denial-of-service attack. Automatic monitoring can for example be implemented using anomaly detection algorithms and machine learning to grade the performance of each executor. Administrators can then be automatically notified if the grade of an executor exceeds a certain threshold, enabling them to take appropriate actions.

\section{Implementation}
Colonies is implemented in Golang and is publicly available on GitHub\footnote{https://github.com/colonyos/colonies} under the MIT license. It consists of a statically compiled binary that offers a CLI tool for deploying Colonies servers or managing the system. Furthermore, there are several software development kits (SDKs) available for various programming languages, including Golang\footnote{https://github.com/colonyos/colonies/tree/main/pkg/client}, Rust\footnote{https://github.com/colonyos/rust}, Julia\footnote{https://github.com/colonyos/Colonies.jl}, JavaScript\footnote{https://github.com/colonyos/colonies.js}, Python\footnote{https://github.com/colonyos/pycolonies}, and Haskell\footnote{https://github.com/colonyos/haskell}. As an introduction to Colonies, the following section gives a brief overview how to use the Python SDK.

@@

The first step in implementing an executor in Python is to create a private key and identity, followed by registering the executor with a colony. As mentioned before, executor can registration can only be carried out by the colony owner using the colony private key. Code in Listening \ref{code:regexecutor} illustrates how to register an executor, including registering functions a specific executor is capable of running. In the example, an executor of type \emph{helloworld\_executor} is registered to run a function named \emph{helloworld}.

@@

After registration, the executor must connect to the Colonies server and request process assignments of the type \emph{helloworld\_executor}. Note that executor properties, such as the executor type, are not explicitly set in the \emph{assign} request, but rather derived implicitly by the Colonies server from the executor identity, which is derived from the signature generated by the executor's private key. This approach ensures that the executor can only be assigned processes that match its registered capabilities.

@@

This \emph{assign} request is always initiated by the executor, allowing executors to be deployed behind firewalls. When receiving the request, the Colonies server hangs the request until a matching process is found or a timer expires. In the example in Listing \ref{code:assign}, the timer is set for 10 seconds, after which an exception is raised, and the executor must reconnect to the server.

When assigned to a process, the executor interprets the metadata stored in the process data structure, including the function name and arguments, and performs some kind of computation such as preprocessing data or training a neural network. Upon completion, the executor closes the process and sets the output. In the example, the output is set to the string \emph{helloworld}.

@@

\section{Appendix - Colonies Operations}
